---
title: "Preliminary CVD cytokine analysis"
author: "Laura Pyle"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(knitr)
library(dplyr)
library(data.table)
library(tidyr)
library(Hmisc)

####################################################################################################
#Data Manipulation #
####################################################################################################

# read main data sheet
all <- read.csv("H:/Endocrinology/Nadeau/T1D Exchange metformin and lipids/Preliminary CVD cytokine data/Jenny manifest v4 GRP2 9th Jan 2019.csv",na.strings = c("","NA"," "))
# delete records with missing data
all <- all[!is.na(all$IFN.g),]

source("C:/Users/pylell/Documents/GitHub/General-code/temp_table1.r")

# read in randomization
rand <- read.csv("H:/Endocrinology/Nadeau/T1D Exchange metformin and lipids/Preliminary CVD cytokine data/metformin vs placebo.csv")

# merge datasets
alldata <- merge(all,rand,by=c("ANALYTICID"))

# get rid of unneeded variables
alldata <- select(alldata,-c("RN","PARENT_ID","SAMPLE_ID","STORAGETYPE","SAMPLEGROUP","BOX.ID","NO.","ROW","COL","pg.ml","X"))

# assign variable for pre vs. post treatment
alldata <- alldata[order(alldata$ANALYTICID,alldata$COLLECTIONDT),]
alldata <- alldata %>% group_by(ANALYTICID) %>% mutate(id=row_number())
alldata$prepost[alldata$id==1] <- "PRE"
alldata$prepost[alldata$id==2] <- "POST"

# Tim deltas
nums <- unlist(lapply(alldata, is.numeric)) 
deltas <- alldata[alldata$prepost == "POST",nums] - alldata[alldata$prepost == "PRE",nums]
deltas$ANALYTICID <- alldata$ANALYTICID[alldata$prepost == "PRE"]
# merge datasets
deltas <- merge(deltas,rand,by=c("ANALYTICID"))

# labels
label(deltas$IFN.g)="Change in IFN-g"
label(deltas$IL.1b)="Change in IL.1b"
label(deltas$IL.12.IL.23p40)="Change in IL.12.IL.23p40"
label(deltas$IL.17A)="Change in IL.17A"
label(deltas$IL.18)="Change in IL.18"
label(deltas$IL.6)="Change in IL.6"
label(deltas$IL.8)="Change in IL.8"
label(deltas$MCP.1..Human.)="Change in MCP.1..Human."
label(deltas$TNF.a)="Change in TNF.a"


# compare deltas in the treatment groups
table <- final_table(deltas,c("IFN.g","IL.1b","IL.12.IL.23p40","IL.17A","IL.18","IL.6","IL.8","MCP.1..Human.","TNF.a"),
                     group=as.factor(deltas$Treatment.Group),margin=2,single=0,ron=2)

N <- nrow(deltas)

```

# Background

The purpose of this analysis is to compare the change in CVD inflammatory markers in participants on metformin to participants on placebo in the T1D Exchange metformin study.

# Methods

The change in each inflammatory marker was calculated as the post-treatment value minus the pre-treatment value.  The distributions of these changes were examined prior to analysis.  For normally distributed variables, the mean and standard deviation were calculated, and groups were compared using t-tests.  For variables that were not normally distributed, the median, 25th and 75th percentiles were calculated, and the groups were compared using the Mann Whitney test.

# Results

There were `r N` subjects included in this analysis.  Changes in inflammatory CVD markers are shown in Table 1.

```{r, results='asis',tidy=TRUE, echo=FALSE}
kable(table,caption="Table 1.  Change in CVD inflammatory markers in metformin and placebo groups.  Descriptive statistics are mean ± standard deviation for normally distributed variables, or median (25th percentile, 75th percentile for variables that were not normally distributed.  Normally distributed variables were compared using t-tests; variables that were not normally distributed were compared using the Mann Whitney test.")
```


